---
title: "Untitled"
author: "Lisa Hopcroft"
date: "14 November 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)

```

# Dynamic specification


```{r data_creation}

set.seed( 5482 )

number_of_patients = 100

trial_arms = c( "Control", "Drug1", "Drug2" )
YN_categories = c( "Yes", "No" )

dummy.d = tibble( 
  patientID = sprintf( "N%03d", 1:number_of_patients ),
  
  arm = sample( trial_arms,
                number_of_patients,
                replace=TRUE),
  
  ### Categorical variables
  has_had_surgery      = sample( YN_categories,
                                 number_of_patients,
                                 replace = TRUE ),
  has_had_radiotherapy = sample( YN_categories,
                                 number_of_patients,
                                 replace = TRUE ),
  has_had_chemotherapy = sample( YN_categories,
                                 number_of_patients,
                                 replace = TRUE ),
  has_had_immunotherapy = sample( YN_categories,
                                 number_of_patients,
                                 replace = TRUE ),
  
  ### Demographics
  age    = round( rnorm( number_of_patients, mean=60, sd=10), 0 ),
  weight = round( rnorm( number_of_patients, mean=70, sd=15), 2 ),
  height = round( rnorm( number_of_patients, mean=170, sd=15), 2 )
  
)

```

## Tabulating data

I need to show how many patients have had surgery in each arm.

```{r}

has_had_surgery.table = dummy.d %>%
  group_by( arm, has_had_surgery ) %>%
  summarise( count=n() )

has_had_surgery.table = dummy.d %>%
  group_by( arm, has_had_surgery ) %>%
  summarise( count=n() ) %>%
  spread( has_had_surgery, count )

```


Now I need to show how many patients have had radiotherapy.
What do I have to change?

```{r}

has_had_surgery.table = dummy.d %>%
  group_by( arm, has_had_surgery ) %>%
  summarise( count=n() ) %>%
  spread( has_had_surgery, count )









have_had_radiotherapy.table = dummy.d %>%
  group_by( arm, has_had_radiotherapy ) %>%
  summarise( count=n() ) %>%
  spread( has_had_radiotherapy, count )

```

What would I have to change to show the number of patients
that have had chemotherapy?

```{r}

has_had_surgery.table = dummy.d %>%
  group_by( arm, has_had_surgery ) %>%
  summarise( count=n() ) %>%
  spread( has_had_surgery, count )









have_had_chemotherapy.table = dummy.d %>%
  group_by( arm, has_had_chemotherapy ) %>%
  summarise( count=n() ) %>%
  spread( has_had_chemotherapy, count )

```


We are coping and pasting.  And also mutliple things are the same.

So, what should we do?  Make a loop.

But how to we refer to things when the tidyverse uses
'Non Standard Evaluation'?  We have to use the "quote/unquote"
process, where *quote* means turning the text into something
that the tidyverse will recognise and *unquote* means inserting
the object into a tidyverse type structure.

This is quote/unquote process is usually done invisibly by
the tidyverse, but here we are cracking it open, looking under
the hood to take advantage of the tidyverse functionality and
use it more flexibly.

There are a few ways of doing this, but let's start with 
using the function `sym()` to *quote* and `!!` to *unquote*
(you can call `!!` bang-bang).

```{r}

column_list = syms( c( "has_had_surgery",
                       "has_had_radiotherapy",
                       "has_had_chemotherapy" ) )

for ( this_column in column_list ) {
  
  table_out = dummy.d %>%
    group_by( arm, !!this_column ) %>%
    summarise( count=n() ) %>%
    spread( !!this_column, count )
  
  print( kable( table_out ) )
}

```

We can also do this for plots.  Here is an example of a plot
we might want to generate for this dataset.

```{r}

ggplot( dummy.d,
        aes( x=arm, y=age )) +
  geom_boxplot() +
  ggtitle( "Boxplot of age across arms" )

```

How can I change this to do the same thing for weight and height?

* *quote* the variables using `sym()`
* *unquote* the variables when using in a tidyverse context using `!!`
* You can interpret the `sym()` object using the `as_name()` function from the `rlang` library.

```{r}

column_list = syms( c( "age",
                       "weight",
                       "height" ) )

for ( this_column in column_list ) {
  plot_out = ggplot( dummy.d,
          aes( x=arm, y=!!this_column )) +
    geom_boxplot( ) +
  ggtitle( sprintf( "Boxplot of %s across arms",
                    rlang::as_name(this_column) ) )
  print( plot_out )
}

```

Now what we _really_ want to do is wrap this up into a function.

Let's write a function.

```{r}

### First create the function
draw_boxplot_for_arms = function( d,
                                  this_column ) {
  
  plot_out = ggplot( d,
          aes( x=arm, y=!!this_column )) +
    geom_boxplot( ) +
  ggtitle( sprintf( "Boxplot of %s across arms",
                    rlang::as_name(this_column) ) )
  return( plot_out )
 
}


### Then insert a call to this function where
### we had the previous code in the loop.
column_list.strings = syms( c( "age",
                       "weight",
                       "height" ) )

for ( this_column in column_list.strings ) {
  p = draw_boxplot_for_arms( d=dummy.d,
                       =this_column )
  print(p)
}

```

Another option: quosures.  Quosures are able to handle strings, like
`sym()`.  But they can also handle the bare column names, as you would
provide to any normal `dplyr` or `ggplot` command.

So, when in doubt, it is best to use `quo` rather than `sym`.

```{r}

### quo() can handle the bare variables
draw_boxplot_for_arms( d          =dummy.d,
                       this_column=quo(age) )

### ...but it can also handle strings like sym()
draw_boxplot_for_arms( d          =dummy.d,
                       this_column=quo("age") )


### We can use the same function as we defined above
### as we can use the !! operator in the same way.
column_list.quosures = quos( age,
                             weight,
                             height )

for ( this_column in column_list.quosures ) {
  p = draw_boxplot_for_arms( d=dummy.d,
                             this_column=this_column )
  print(p)
}

```

In the previous example, we were passing a quosure to the function.
We could also convert to the required quosure within the function.
For this, we need to use the `enquo()` function.

NB. Plural forms:
sym -> syms
quo -> quos
enquo -> enquos

Note: I don't know why quo doesn't work within a function.

```{r}


draw_boxplot_for_arms_enquo = function( d,
                                        this_column ) {
  
  plot_out = ggplot( d,
          aes( x=arm, y=!!enquo(this_column) )) +
    geom_boxplot( ) +
  ggtitle( sprintf( "Boxplot of %s across arms",
                    rlang::as_name(enquo(this_column)) ) )
  return( plot_out )
 
}

draw_boxplot_for_arms_enquo( d          =dummy.d,
                             this_column=age)



```

So, up until now, we have been converting our column name into an
object that the tidyeval understands (using either `sym()` or `quo()`
depending on whether you are starting with a string or bare
representation of your column name respectively) and then using the
`!!` to integrate that into some tidyverse code (`ggplot()` or
`dplyr` functions).

In the `rlang` package v0.4.0 and above, you can do the quote
and the unquote *at the same time* (in one 'interpolation' step),
using the "curly curly" or "superstache" operator.

```{r}

draw_boxplot_for_arms_curlycurly = function( d,
                                             this_column ) {
  
  plot_out = ggplot( d,
          aes( x=arm, y={{this_column}} )) +
    geom_boxplot( ) +
  ggtitle( sprintf( "Boxplot of %s across arms",
                    rlang::as_name(enquo(this_column)))  )
  
  return( plot_out )
 
}

draw_boxplot_for_arms_curlycurly( d=dummy.d,
                                  this_column=age )


```

One last thing: what if I want to add a new column and the name of that
column is tied up inside a variable?

```{r}

### Doing this with bare variable names
dummy.target.d_bare = dummy.d %>%
  mutate( chemo_but_no_radio =
            ( has_had_chemotherapy == "Yes" & 
            has_had_radiotherapy == "No" ) )
nrow( dummy.target.d_bare )
dummy.target.d_bare %>% colnames

### What if the new column name is saved
### as a string in a variable?

new_variable = "chemo_but_no_radio"

dummy.target.d_byVar1 = dummy.d %>%
  mutate( !!sym(new_variable) := ( has_had_chemotherapy  == "Yes" & has_had_radiotherapy == "No" ) )

dummy.target.d_byVar2a = dummy.d %>%
  mutate( !!quo(new_variable) := ( has_had_chemotherapy  == "Yes" & has_had_radiotherapy == "No" ) )

dummy.target.d_byVar2b = dummy.d %>%
  mutate( !!enquo(new_variable) := ( has_had_chemotherapy  == "Yes" & has_had_radiotherapy == "No" ) )

dummy.target.d_byVar3 = dummy.d %>%
  mutate( {{new_variable}} := ( has_had_chemotherapy  == "Yes" & has_had_radiotherapy == "No" ) )

new_variable %in% ( dummy.target.d_byVar1  %>% colnames )
new_variable %in% ( dummy.target.d_byVar2a %>% colnames )
new_variable %in% ( dummy.target.d_byVar2b %>% colnames )
new_variable %in% ( dummy.target.d_byVar3  %>% colnames )

```


Useful references for more reading:

* https://dplyr.tidyverse.org/articles/programming.html
* https://tidyeval.tidyverse.org/sec-why-how.html
* https://edwinth.github.io/blog/dplyr-recipes/
* https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/
* https://stackoverflow.com/questions/56936372/curly-curly-tidy-evaluation-and-modifying-inputs-or-their-names
* https://www.tidyverse.org/blog/2018/07/ggplot2-tidy-evaluation/
* https://rlang.r-lib.org/reference/quasiquotation.html

For more information and to really understand how this works, see:

https://adv-r.hadley.nz/quasiquotation.html




